Uniform formula lower bounds for testing the existence of and counting satisfiable assignments

### 2022-06-24 article/main.tex
intro almost done

\newcommand{\Poly}{\textsf{P}}%
\newcommand{\Log}{\textsf{L}}%
\newcommand{\NP}{\textsf{NP}}%
\newcommand{\TIME}{\textsf{TIME}}%
\newcommand{\NTIME}{\textsf{NTIME}}%
\newcommand{\coNTIME}{\textsf{coNTIME}}%
\newcommand{\SAT}{\textsf{SAT}}%
\newcommand{\ParSAT}{\ensuremath\bigoplus\textsf{SAT}}%
\newcommand{\SharpSAT}{\#\textsf{SAT}}%
\newcommand{\NAND}{\textsf{NAND}}%
\newcommand{\ND}{\textsf{NDepth}}%
\newcommand{\NDL}[1]{\ND[ #1 \log n]}%

\newcommand{\todohere}{\todo[inline]{TODO}}%

@@

Motivation: understanding complexity. Explicit lower bounds : programs, circuits, formulas.
Uniformity.

A line of work initiated by Fortnow~\cite{fortnow2000time,fortnow2005time} in 1997 
has proven model-independent time-space lower bounds for the SAT
problem and related problems within the polynomial-time hierarchy.
This approach was then extended by Williams~\cite{williams2006inductive,williams2007time,williams2013alternation} 
into a complex framework which culminated in proving that \SAT{} 
does not have a small-space algorithm that runs in time $n^{2\cos(\pi/7)} \simeq n^{1.801}$.
This line of work concluded~\cite{buss2015limits} with a proof 
that the $2\cos(\pi/7)$ exponent is optimal for this framework,
which implies that radically new ideas would be required to separate \NP{} from \Log{}.
This approach was later expanded to obtain slightly stronger lower bounds
for quantum classes, and against small-space randomized computation~\cite{mudigonda2020time}.

In this work, we explore a new extension of this line of work, 
which gives lower bounds on the size uniform formulas that solve \SAT{}
and related problems, such as \ParSAT{} and \SharpSAT{}.
By giving an efficient simulation of uniform log-depth \NAND{} formulas 
by $\Sigma_2$ computation, we are able to show that \SAT{}
does not have \textit{uniform} \NAND{} formulas 
of depth less than~$4 \cos(\pi/7) \log n \simeq 3.603 \log n$.

TODO: We also extend this result to show that \SharpSAT{} does not have
alternating AND-OR/XOR-EQ formulas of depth less than~$3.603 \log n$.
\todo[inline]{does that extend to $n^c$-size lower bound for some $c > 3$?}

Finding strong lower bounds on the formula size of explicit functions in \NP{}
is a long standing open problem.
The best lower bounds known depend on the computational basis, that is, 
the set of gates that are allowed in the formula.
For the full binary basis $B_2$, Nechiporuk\cite{nechiporuk66boolean}
proved in 1966 an $\Omegat{n^2}$ lower bound for an explicit function in \NP{}, 
which remains unbeaten to this day.
For the De Morgan basis $\{\wedge, \vee, \neg\}$, Andreev~\cite{andreev1987method} 
proved an $\Omegat{n^{2.5}}$ lower bound, 
which was later improved by H{\aa}stad~\cite{hastad1998shrinkage} to $\Omegat{n^{3}}$.
This lower bound is optimal up to polylog factors,
since their function has (uniform) formulas of size $\Ot{n^{3}}$.
% Another interesting case is that of mononotone formulas, i.e. formulas over the $\{\wedge, \vee\}$ basis,
% for which Razborov, Alon and Boppana~\cite{razborov1985lower, alon1987monotone} showed
% a superpolynomial $n^{\Theta(\log n)}$ lower bounds for the \NP-complete \textsf{Clique} problem.
\todo{add monotone circuits lower bounds? But that's not formulas}

Note that all the lower bounds mentioned in the previous paragraph 
are \textit{size} lower bounds of the form $n^c$,
and they hold for \textit{non-uniform} formulas.  
The lower bounds that we give are \textit{depth} lower bounds of the form $c\log n$,
and only hold for \textit{uniform} formulas.
While a $n^c$ size lower bound implies a $c\log n$ depth lower bounds,
the converse is not immediately true; 
one may get an $n^c$ size lower bound from an
$\alpha c\log n$ depth lower bound, where $\alpha$ depends on the computation basis,
using the work of Brent~\cite{brent1974parallel} and Spira~\cite{spira1971time}.
\todo{can we show something here? e.g. improve for $B_2$?}

@@

We first study \NAND{} formulas. 
The \NAND{} basis is rather expressive: Andreev's function has \NAND{} formulas of depth $(3+o(1)) \log n$ (and therefore of size $\Ot{n^3}$), which matches the lower bound for the De Morgan Basis.

Namely, we prove the following:
\begin{theorem}
	For any $c < 4 \cos(\pi/7) \simeq 3.603$, \SAT{} does not have \NAND{} formulas of depth $(c + o(1)) \log n$.
\end{theorem}

@@

\subsection{Uniform formulas}
A family of formulas

In this work, we focus on formulas with depth $d(n) = (c + o(1))\log n$, for some $c > 1$.

\begin{definition}[``Usual'' polylog-time uniform formulas]
	A family of formulas $(\varphi_n)_{n \ge 1}$ of depth $(c + o(1))\log n$ 
	is \textit{polylog-time uniform} if there exists a machine $A_\varphi$
	such that for every $n$ and every $b\in\bit^{\leq (c + o(1))\log n}$,
	$A_\varphi(b, n)$ outputs a description of the gate at position $b$ in $\varphi_n$ in time $\log^{O(1)} n$.
\end{definition}

Here, the position of a gate is given by a bitstring of length at most $(c + o(1))\log n$ 
that describes the path to follow to reach that gate from the root, 
with $0$ meaning ``go to left child'' and $1$ ``go to right child''.

In this work, we consider a stronger model of uniform formulas, 
where the descriptor machine $A_\varphi$ has random access to the input $x$ 
at which we wish to evaluate the formula, but is still limited to polylogarithmic time.
\begin{definition}[Input-aware polylog-time uniform formulas]\todo{is this a good name?}\label{def:unif}
	A family of formulas $(\varphi_n)_{n \ge 1}$ of depth $(c + o(1))\log n$ is \textit{polylog-time uniform}
	if there exists a \textbf{RAM} machine $B_\varphi$
	such that for every $n$, every $x\in\bit^n$ and every 
	$b\in\bit^{\leq (c + o(1))\log n}$,
	$B_\varphi(x, b, n)$ outputs a description of the gate at position $b$ in $\varphi_n$ in time $\log^{O(1)} n$.
\end{definition}
Notice that this model supersedes the previous one: 
$A_\varphi$ satisfies the conditions of Definition~\ref{def:unif}, 
but does not use its random access to the input, 
therefore the lower bounds that we give against model also hold for the usual one.

Moreover, this definition allows for an efficient simulation of uniform formulas by RAM machines, 
due to the negligible overhead.
Indeed, for every $c > 1$, we have $\ND[c \log n] \subseteq TS[n^c]$.

We will see the need for this stronger definition in \todo{ref}, 
where we need to give $O(\log n)$ bits of information to the descriptor
to adaptively choose a subformula.

Uniform \NAND{}/alternating formulas. Specific model, comparison with the usual model.

@@

We now introduce ``quantified'' complexity classes, which extend alternating classes 
by adding a ``parity'' quantifier.

Given a complexity class $\Cc$ and constants $k, (a_i)_{i\leq k}$, 
we consider \textit{quantified} complexity
class $(Q_1~n^{a_1})\ldots(Q_k~n^{a_k})~\Cc$, 
where $Q_i$ is one of $\forall$ or $\exists$
 % or $\oplus$.
A computation in a quantified complexity class is defined as follows:
\begin{definition}[Quantified computation]
	A language $L$ is in $(Q_1~n^{a_1})\ldots(Q_k~n^{a_k})~\Cc$ 
	is defined by $k$ RAM machines $A_1, \ldots, A_k$ and a $\Cc$ machine $M$
	such that, given an input $x$ of length $n$:
	for every $i \le k$, $A_{i}$ takes as input $x_i$ along with $y_i$, 
	a string of $n^{a_{i} + o(1)}$ nondeterministic bits,
	runs in time $n^{a_{i} + o(1)}$ and outputs a string $x_{i}$ of length at most $n^{a_{i} + o(1)}$, 
	with the convention that $x_0 := x$.
	The output of the last stage, $x_{k}$, is passed to the machine $M$.

	The computation accepts $x$ \textit{starting from stage $i$} if either:
	\begin{itemize}
		\item $i = k+1$ and $M$ accepts $x_{k}$, 
		\item $Q_i$ is $\exists$, and there exists a $y_i$ such that the computation
		accepts $x_{i}$ starting from stage $i+1$, 
		\item $Q_i$ is $\forall$, and for every $y_i$, the computation
		accepts $x_{i}$ starting from stage $i+1$. 
		% \item $Q_i$ is $\bigoplus$, the number of $y_i$ such that the computation
		% accepts $x_{i}$ starting from stage $i+1$ is odd. 
	\end{itemize}

	An input $x$ is in $L$ if the computation $x$ accepts starting from stage 1.
\end{definition}
\todo[inline]{allow $O(\log n)$ quantifiers ?}

\subsection{Overview of the methodology}

The general methodology used in this paper to obtain lower bounds
extends the one used in~\cite{mudigonda2020time}.

To prove that \SAT{} does not have formulas of depth less than $c\log n$, 
we reason by contradiction.
First, we assume that \SAT{} has such formulas, and the strong \NP-completeness of \SAT{}
allows us to deduce that the same is holds for all of $\NTIME[n]$, 
\begin{lemma}\label{lemma:usual_sd}
	If \SAT{} is in $\NDL{c}$ for some $c>0$,
	then $\NTIME[n] \cup \coNTIME[n] \subseteq \NDL{c}$.
\end{lemma}

We also show how to speed-up (or rather, reduce the depth of) computations by adding quantifiers, 
using various ``Speedup rules''. 
For example, for NAND formulas (which are the same as alternating AND/OR formulas), we have the following:
\begin{lemma}[Speedup rule]
	For every $c > 0, x \leq c$, we have:
	\[\NDL{c} \subseteq (\exists n^{x/2}) (\forall n^1) \NDL{(c-x)}\]
\end{lemma}

The above lemma can be combined with exhaustive search and Lemma~\ref{lemma:usual_sd} 
to yield an improved version of Lemma~\ref{lemma:usual_sd} for quantified classes, 
which we call the ``Slowdown rule''.
\begin{lemma}[Slowdown rule]
	Assume that \SAT{} is in $\NDL{c}$ for some $c>0$.
	Then, for every $a,b,d > 0$, we have:
	\[\ldots (Q~n^{a}) (\neg Q~n^b) \NDL{d}
	\subseteq \ldots (Q~n^{a}) \NDL{c\cdot\max(d/2, a, b)}.\]
	% \[\ldots (Q_{k-1}~n^{a_{k-1}}) (Q_k~n^{a_k}) \NDL{d} 
	% \subseteq (Q_{k-1}~n^{a_{k-1}}) \NDL{c\cdot \max(d/2, a_k, a_{k-1})}\]	
\end{lemma}

@@

%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%
\section{Proof of main result}

\subsection{Speedup and Slowdowns}
We start by discussion the Speedup rule for Alternating formulas.

\begin{lemma}[Speedup rule]\label{lemma:speedup}
	For every $d > 0$ and $0 \ge x \ge d$, we have
	\begin{flalign*}
	\NDL{d} &\subseteq (\exists n^{x/2}) (\forall \frac{x}{2}\log n) \NDL{d-x}\text{ and}\\
	\NDL{d} &\subseteq (\forall n^{x/2}) (\exists \frac{x}{2}\log n) \NDL{d-x}
	\end{flalign*}
\end{lemma}
\todo[inline]{fix quantifiers?}

@@

We can then eliminate OR gates and replace them with their selected child.
We are left with a formula starting with $(x/2)\log n$ levels of AND gates, 
followed by smaller OR/AND formulas of depth $(d-x + o(1)) \log n$.
The main formula evaluate to 1 if and only if for every of these $O(n^{x/2})$
smaller formulas evaluates to true: 
we can check this using a universal quantifiers, 
which requires $O(\frac{x}{2}\log n)$ bits, and then running an $\NDL{d-x}$ computation.

@@

\begin{proof}[Proof of Lemma~\ref{lemma:speedup}]
	We prove the result for $\exists\forall$ computations, 
	the result for $\forall\exists$ computations follows from the fact
	that $\NDL{d}$ is closed under complementation.

@@

	We describe how to build bit strings $y$ of length $n^{x/2}$ and
	\todohere 
\end{proof}

\begin{corollary}\label{cor:speedup}
	For every $d > 0$, we have
	\[\NDL{d} \subseteq \NTIME[n^{d/2}] \cap \coNTIME[n^{d/2}].\]
\end{corollary}
\begin{proof}
	Applying Lemma~\ref{lemma:speedup} with $x = d$, we obtain
	$\NDL{d} \subseteq (\exists n^{d/2}) (\forall \frac{d}{2}\log n) \NDL{0}$.
	Now, using an exhaustive search over the $n^{d/2}$ bit strings of length $\frac{d}{2}\log n$
	followed by a deterministic $n^{o(1)}$ computation, we get that
	$\NDL{d} \subseteq (\exists n^{d/2}) \TIME[n^{d/2}] = \NTIME[n^{d/2}]$.

@@

We now discuss how to efficiently remove quantifiers from an alternating \ND{} computation.
Slowdown rule from algorithms for SAT assumptions:
\begin{lemma}[Slowdown rule]\label{lemma:slowdown}
	If $\SAT{}\in \NDL{c}$, then for every $d \ge 1$, $\NTIME[n^d] \cup \coNTIME[n^d] \subseteq \NDL{c\cdot d}$.
\end{lemma}

A key idea for the proof of Lemma~\ref{lemma:slowdown} 
is the fact that any language in $\NTIME[n]$ can
be reduced to $\SAT{}$ via uniform poly-log time first-order projections, 
with instances of size $n$ reduced to formulas of size $n^{1+o(1)}$.
Intuitively, a first order projection is a depth 0 circuit, 
i.e. each output bit is either a constant (0 or 1),
an input $x_i$ or its negation, $\neg x_i$.
For a more in-depth exposition of first-order projections,
see for example~\cite[End of Sec.~3]{allender1997first}.
\begin{proof}[Proof of Lemma~\ref{lemma:slowdown}]
	We first prove that $\SAT{}\in \NDL{c}$ implies that $\NTIME[n] \subseteq \NDL{c}$,
	then use a padding argument to show that $\forall d\ge 1, \NTIME[n^d] \subseteq \NDL{c\cdot d}$,
	and the result follows since $\NDL{c\cdot d}$ is closed under complement.

	Let $L\in\NTIME[n]$. 
	As stated above, $L$ is Karp-reducible to \SAT{} via
	uniform poly-log time first-order projections, 
	\todo{ref for this? Folklore?}
	with instances of size $n$ reduced to formulas of size $n^{1+o(1)}$.
	Now, notice that the composition of a NAND formula of depth $d$ with a first order projection
	is also a NAND formula of depth $d$: it only consists in relabeling leaves of the formula.
	Poly-log time uniformity is also preserved. 
	Therefore, we can use $(c+o(1)) \log n$-depth NAND formulas for \SAT{} on $n^{1+o(1)}$
	inputs to decide $L$ by composing them with the first order projection; 
	the resulting formula has size $(c+o(1)) \log n^{1+o(1)} = (c+o(1)) \log n$,
	therefore $L\in\NDL{c}$.

	We now use a padding argument to lift this result to $\NTIME[n^d]$.
	Let $d>1$, let $L\in \NTIME[n^d]$ and let $M$ be a nondeterministic machine that decides $L$
	in time $n^{d+o(1)}$.
	Consider the padded language $L' = \{x1^{|x|^d - |x|} \mid x\in L\}$.
	Given an input $y$ of length $m$, we can decide whether $y\in L'$ in time 
	$m^{1+o(1)}$ as follows:
	\begin{itemize}
		\item if $y$ is not of the form $x1^{|x|^d - |x|}$ for some $x$, reject $y$.
		This can be done in time $O(m)$.
		\item otherwise, $y = x1^{|x|^d - |x|}$. Run $M$ on $x$, and accept $y$ 
		if and only it accepts $x$. This takes time $|x|^{d+o(1)} = m^{1+o(1)}$.
	\end{itemize}
	This implies $L'$ is in $\NTIME[n]$, which in turn implies that $L'\in\NDL{c}$.
	We can then use the formula descriptor $A$ for $L'$ 
	to describe formulas of depth $(c + o(1))\log \left(n^{d+o(1)}\right) = (c\cdot d + o(1))\log n$ for $L$.
	On input $x,b,n$, the descriptor machine $B$ for $L$  
	calls $A$ on $x, b, n^d$.
	If $A$ output a gate type or a constant, $B$ outputs the same information.
	If $A$ outputs a literal $x_i^*$, $B$ outputs $x_i^*$ if $i \le n$, and 1 otherwise.
\end{proof}

@@

%%%%%%%%%%%%%%%%%%%%%%
\subsection{Structure of optimal proofs}
How to get the contradiction given the slowdown rule for $c < 4\cos(\pi /7)$.

>>>>

\newcommand{\Poly}{\textsf{P}}%
\newcommand{\Log}{\textsf{L}}%
\newcommand{\NP}{\textsf{NP}}%
\newcommand{\TIME}{\textsf{TIME}}%
\newcommand{\NTIME}{\textsf{NTIME}}%
\newcommand{\coNTIME}{\textsf{coNTIME}}%
\newcommand{\SAT}{\textsf{SAT}}%
\newcommand{\ParSAT}{\ensuremath\bigoplus\textsf{SAT}}%
\newcommand{\SharpSAT}{\#\textsf{SAT}}%
\newcommand{\NAND}{\textsf{NAND}}%
\newcommand{\ND}{\textsf{NDepth}}%
\newcommand{\NDL}[1]{\ND\left[ #1 \log n\right]}%
\newcommand{\TS}[1]{\textsf{TS}\left[ n^{ #1 }\right]}%

\newcommand{\todohere}{\todo[inline]{TODO}}%
\newcommand{\todoh}[1]{\todo[inline]{TODO: #1 }}%

@@

\todoh{Motivation: understanding complexity. Explicit lower bounds: programs, circuits, formulas.
Uniformity.}

A line of work initiated by Fortnow~\cite{fortnow2000time,fortnow2005time} in 1997 
has proven model-independent time-space lower bounds for the SAT
problem and related problems within the polynomial-time hierarchy.
This approach was then extended by Williams~\cite{williams2006inductive,williams2007time,williams2013alternation} 
into a complex framework which culminated in proving that \SAT{} 
does not have a small-space algorithm that runs in time $n^{2\cos(\pi/7)}  \simeq n^{1.801}$.
This line of work concluded~\cite{buss2015limits} with a proof 
that the $2\cos(\pi/7)$ exponent is optimal for this framework,
which implies that radically new ideas would be required to separate \NP{} from \Log{}.
This approach was later extended to obtain slightly stronger lower bounds
for quantum classes, and against small-space randomized computation~\cite{mudigonda2020time}.

In this work, we explore a new extension of this line of work, 
which gives lower bounds on the depth of uniform formulas that solve \SAT{}.
By giving a new efficient simulation of uniform log-depth \NAND{} formulas 
by $\Sigma_2$ computation, we are able to show that \SAT{}
does not have \textit{uniform} \NAND{} formulas 
of depth less than~$4 \cos(\pi/7) \log n \simeq 3.603 \log n$.

Finding strong lower bounds on the formula size of explicit functions in \NP{}
is a long-standing open problem.
The best lower bounds known depend on the computational basis, that is, 
the set of gates that are allowed in the formula.
For the full binary basis $B_2$, Nechiporuk\cite{nechiporuk66boolean}
proved in 1966 an $\Omegat{n^2}$ lower bound for an explicit function in \NP{}, 
which remains unbeaten to this day.
For the De Morgan basis $\{\wedge, \vee, \neg\}$, Andreev~\cite{andreev1987method} 
proved an $\Omegat{n^{2.5}}$ lower bound, 
which was later improved by H{\aa}stad~\cite{hastad1998shrinkage} to $\Omegat{n^{3}}$.
This lower bound is optimal up to polylog factors,
since their function has (uniform) formulas of size $\Ot{n^{3}}$.

Note that all the lower bounds mentioned in the previous paragraph 
are \textit{size} lower bounds of the form $n^{c + o(1)}$
and they hold for \textit{non-uniform} formulas.  
The lower bounds that we give are \textit{depth} lower bounds of the form $(c + o(1))\log n$,
and only hold for \textit{uniform} formulas.
While an $n^c$ size lower bound implies a $c\log n$ depth lower bounds,
the converse is not immediately true; 
one may get an $n^c$ size lower bound from an
$\alpha c\log n$ depth lower bound, where $\alpha$ depends on the computation basis,
using the work of Brent~\cite{brent1974parallel} and Spira~\cite{spira1971time}.

@@

The \NAND{} basis is rather expressive: Andreev's function has \NAND{} formulas of depth $(3+o(1)) \log n$ (and therefore of size $\Ot{n^3}$), which matches the lower bound for the De Morgan Basis.
\todoh{more about nand?}

Namely, we prove the following:
\begin{theorem}\label{thm:main}
	For any $c < 4 \cos(\pi/7) \simeq 3.603$, \SAT{} does not have 
	uniform \NAND{} formulas of depth $(c + o(1)) \log n$.
\end{theorem}

@@

\subsection{Uniform formulas}
A boolean formula $\phi$ over $n$ inputs is a 
rooted full binary tree\footnote{\textit{Full} means that every node has either two or zero children.}
whose internal nodes, called gates, are labeled by a function $f: \bit^2 \rightarrow \bit$,
and leaves are labeled with a variable, $x_i$ its negation $\neg x_i$ for some $i, 1\le i \le n$,
or a constant $c\in\bit$.
The depth of a formula is the depth of the underlying tree.

In this work, we focus on \NAND{} formulas, that is formulas whose gates are all labeled 
with the function $\NAND: x,y \mapsto \neg(x\wedge y)$.
The De Morgan law implies that $\NAND(x, y) = \neg x \wedge \neg y$, 
therefore \NAND{} formulas are equivalent to ``alternating OR/AND'' formulas (up to negating some inputs), 
that is formulas whose gate with even-depth \footnote{The root has depth 0.} 
are OR gates, and odd-depth gates are AND gates.
This alternative alternating point of view will be useful 
when describing a procedure to compute the value of a formula with alternating computation.

\todoh{To introduce: family of formulas computing a language}
% In this work, we focus on formulas with depth $d(n) = (c + o(1))\log n$, for some $c > 1$.
However, there exists formulas of \textit{constant} depth that compute undecidable languages.
In this work, we restrict our attention to uniform formulas, 
that is family of formulas for which the description of the
$n$-input formula can be computed efficiently.
We first recall a usual definition of uniform formulas.
\begin{definition}[``Usual'' polylog-time uniform formulas]
	A family of formulas $(\varphi_n)_{n \ge 1}$ of depth $(c + o(1))\log n$ 
	is \textit{polylog-time uniform} if there exists a machine $A_\varphi$
	such that for every $n$ and every $b\in\bit^{\leq (c + o(1))\log n}$,
	$A_\varphi(b, n)$ outputs a description of the gate at position $b$ in $\varphi_n$ in time $\log^{O(1)} n$.
\end{definition}

Here, the position of a gate is given by a bit string of length at most $(c + o(1))\log n$ 
that describes the path to follow to reach that gate from the root, 
with $0$ meaning ``go to the left child'' and $1$ ``go to the right child''.

However, in out work, we need to use a stronger model of uniform formulas, 
where the descriptor machine $A_\varphi$ has random access to the input $x$ 
at which we wish to evaluate the formula but is still limited to polylogarithmic time.
\begin{definition}[Input-aware polylog-time uniform formulas]\label{def:unif}
	A family of formulas $(\varphi_n)_{n \ge 1}$ of depth $(c + o(1))\log n$ is \textit{polylog-time uniform}
	if there exists a \textbf{RAM} machine $B_\varphi$
	such that for every $n$, every $x\in\bit^n$ and every 
	$b\in\bit^{\leq (c + o(1))\log n}$,
	$B_\varphi(x, b, n)$ outputs a description of the gate at position $b$ in $\varphi_n$ in time $\log^{O(1)} n$.
\end{definition}
Notice that this model supersedes the previous one: 
$A_\varphi$ satisfies the conditions of Definition~\ref{def:unif}, 
but does not use its random access to the input, 
therefore the lower bounds that we give against this model also hold for the usual one.

We will see the need for this stronger definition in Section~\ref{sec:rules}, 
where we need to give $O(\log n)$ bits of information to the descriptor
to adaptively choose a subformula.

Finally, for any $c \ge 0$, we define $\NDL{c}$
to be the set of languages that have uniform 
input-aware polylog-time NAND formulas of depth$(c+o(1)) \log n$.

Notice that uniform log-depth NAND formulas
can be efficiently simulated by RAM machines, 
due to the negligible runtime overhead of the descriptor.
Indeed, for every $c > 1$, we have $\NDL{c} \subseteq TS[n^c]$.

@@

We now introduce alternating complexity classes, a fine-grained version of the polynomial hierarchy.

Given a complexity class $\Cc$ and constants $k, (a_i)_{i\leq k}$, 
we consider \textit{alternating} complexity
class $(Q_1~n^{a_1})\ldots(Q_k~n^{a_k})~\Cc$, 
where $Q_i$ is one of $\forall$ or $\exists$
A computation in a alternating complexity class is defined as follows:
\begin{definition}[Alternating computation]
	A language $L$ is in $(Q_1~n^{a_1})\ldots(Q_k~n^{a_k})~\Cc$ 
	is defined by $k$ RAM machines $A_1, \ldots, A_k$ and a $\Cc$ machine $M$
	such that, given an input $x$ of length $n$:
	for every $i \le k$, $A_{i}$ takes as input $x_i$ along with $y_i$, 
	a string of $n^{a_{i} + o(1)}$ nondeterministic bits,
	runs in time $n^{a_{i} + o(1)}$ and outputs a string $x_{i}$ of length at most $n^{a_{i} + o(1)}$, 
	with the convention that $x_0 := x$.
	The output of the last stage, $x_{k}$, is passed to the machine $M$.

	The computation accepts $x$ \textit{starting from stage $i$} if either:
	\begin{itemize}
		\item $i = k+1$ and $M$ accepts $x_{k}$, 
		\item $Q_i$ is $\exists$, and there exists a $y_i$ such that the computation
		accepts $x_{i}$ starting from stage $i+1$, 
		\item $Q_i$ is $\forall$, and for every $y_i$, the computation
		accepts $x_{i}$ starting from stage $i+1$. 
	\end{itemize}

	An input $x$ is in $L$ if the computation $x$ accepts starting from stage 1.
\end{definition}

\subsection{Overview of the methodology}\label{sec:overview}

The general methodology used in this paper to obtain lower bounds
extends the one used in~\cite{williams2006inductive, mudigonda2020time}
to the specific case of uniform \textsf{NAND} formulas of logarithmic depth.

To show that \SAT{} does not have formulas of depth less than $c\log n$, 
we reason by contradiction.
First, we assume that \SAT{} has such formulas, and the strong \NP-completeness of \SAT{}
allows us to deduce that the same holds for all of $\NTIME[n]$, 
\begin{lemma}\label{lemma:usual_sd}
	If \SAT{} is in $\NDL{c}$ for some $c>0$,
	then $\NTIME[n] \cup \coNTIME[n] \subseteq \NDL{c}$.
\end{lemma}

We also show how to speed up (or rather, reduce the depth of) computations by adding quantifiers, 
using various ``Speedup rules''. 
For example, for NAND formulas (which are the same as alternating AND/OR formulas), we have the following:
\begin{lemma}[Speedup rule]
	For every $c > 0, x \leq c$, we have:
	\[\NDL{c} \subseteq (\exists n^{x/2}) (\forall n^1) \NDL{(c-x)}\]
\end{lemma}

The above lemma can be combined with exhaustive search and Lemma~\ref{lemma:usual_sd} 
to yield an improved version of Lemma~\ref{lemma:usual_sd} for quantified classes, 
which we call the ``Slowdown rule''.
\begin{lemma}[Slowdown rule]
	Assume that \SAT{} is in $\NDL{c}$ for some $c>0$.
	Then, for every $a,b,d > 0$, we have:
	\[\ldots (Q~n^{a}) (\neg Q~n^b) \NDL{d}
	\subseteq \ldots (Q~n^{a}) \NDL{c\cdot\max(d/2, a, b)}.\]
\end{lemma}

@@

\paragraph{Optimality of the result}
In the case of \SAT{} vs small-space algorithms, 
Buss and Williams~\cite{buss2015limits} proved that the result $\SAT\notin \TS{n^c}$
for $c < 2\cos(\pi/7)$ is optimal for their framework. 
It was later extended and simplified by Mudigonda and Williams~\cite{mudigonda2020time},
who showed upper bounds on .
It appears that their generalization covers our framework,
therefore we cannot prove $\SAT\notin\NDL{c}$ for $c \ge 4\cos(\pi/7)$
without finding a new (and improved) way to speed up \ND{} computation using alternations.

%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%
\section{Proof of the main result}

\subsection{Speedup and Slowdowns}\label{sec:rules}
We start by discussing the Speedup rule for Alternating formulas.

\begin{lemma}[Speedup rule]\label{lemma:speedup}
	For every $d > 0$ and $0 \ge x \ge d$, we have
	\begin{flalign*}
	\NDL{d} &\subseteq (\exists n^{x/2}) (\forall n) \NDL{d-x}\text{ and}\\
	\NDL{d} &\subseteq (\forall n^{x/2}) (\exists n) \NDL{d-x}
	\end{flalign*}
\end{lemma}

@@

We can then eliminate OR gates and replace them with their selected child.
We are left with a formula starting with $(x/2)\log n$ levels of AND gates, 
followed by smaller OR/AND formulas of depth $(d-x + o(1)) \log n$.
The main formula evaluates to 1 if and only if for every of these $O(n^{x/2})$
smaller formulas evaluates to true: 
we can check this using a universal quantifier, 
which requires $O(\frac{x}{2}\log n)$ bits, and then running an $\NDL{d-x}$ computation.
Note that while the second quantifier only requires $O(\frac{x}{2}\log n)$ nondeterministic bits,
it still computes and outputs $n^{1+o(1)}$ bits, hence the $n^1$ exponent.

@@

\begin{proof}[Proof of Lemma~\ref{lemma:speedup}]
	We prove the result for $\exists\forall$ computations, 
	the result for $\forall\exists$ computations follows from the fact
	that $\NDL{d}$ is closed under taking the complement.

@@

	We describe how to build bit strings $y$ of length $n^{x/2}$ and
	\todoh{finish details of proof}
\end{proof}

Lemma~\ref{lemma:speedup} yields the following corollary,
which gives us a simulation of formulas by Turing machines that is more efficient
than the naive evaluation.
\begin{corollary}\label{cor:speedup}
	For every $d > 0$, we have
	\[\NDL{d} \subseteq \NTIME[n^{d/2}] \cap \coNTIME[n^{d/2}].\]
\end{corollary}
\begin{proof}
	Applying Lemma~\ref{lemma:speedup} with $x = d$, we obtain
	$\NDL{d} \subseteq (\exists n^{d/2}) (\forall n) \NDL{0}$.
	However, looking closely at the proof of Lemma~\ref{lemma:speedup}
	reveals that the $\forall$ quantifier here only uses $O(\frac{d}{2}\log n)$
	nondeterministic bits.
	Therefore, we can use exhaustive search over the $n^{d/2}$ bit strings of length $\frac{d}{2}\log n$
	to remove that quantifier.
	This leaves us we a deterministic $n^{o(1)}$ computation, and we get that
	$\NDL{d} \subseteq (\exists n^{d/2}) \TIME[n^{d/2}] = \NTIME[n^{d/2}]$.

@@

We now discuss how to use this corollary to
efficiently remove quantifiers from an alternating \ND{} computation.
The first step is to show that the assumption $\SAT\in\NDL{c}$
propagates efficiently to all of \NP{}.
This result is known as the Slowdown rule: it allows us to remove a quantifier,
at the cost of an increased depth of the verifier.
\begin{lemma}[Slowdown rule]\label{lemma:slowdown}
	If $\SAT{}\in \NDL{c}$, then for every $d \ge 1$, $\NTIME[n^d] \cup \coNTIME[n^d] \subseteq \NDL{c\cdot d}$.
\end{lemma}

A key idea for the proof of Lemma~\ref{lemma:slowdown} 
is the fact that any language in $\NTIME[n]$ can
be reduced to $\SAT{}$ via uniform poly-log time first-order projections, 
with instances of size $n$ reduced to formulas of size $n^{1+o(1)}$.
Intuitively, a first-order projection is a depth 0 circuit, 
i.e. each output bit is either a constant (0 or 1),
an input $x_i$, or its negation $\neg x_i$.
For a more in-depth exposition of first-order projections,
see for example~\cite[End of Sec.~3]{allender1997first}.
\begin{proof}[Proof of Lemma~\ref{lemma:slowdown}]
	We first prove that $\SAT{}\in \NDL{c}$ implies that $\NTIME[n] \subseteq \NDL{c}$
	and use a padding argument to show that 
	\[\forall d\ge 1, \NTIME[n^d] \subseteq \NDL{c\cdot d}.\]
	The result follows since $\NDL{c\cdot d}$ is closed under complement.

	Let $L\in\NTIME[n]$. 
	As stated above, $L$ is Karp-reducible to \SAT{} via
	uniform poly-log time first-order projections, 
	with instances of size $n$ reduced to formulas of size $n^{1+o(1)}$.
	Now, notice that the composition of a NAND formula of depth $d$ with a first-order projection
	is also a NAND formula of depth $d$: it only consists in relabeling leaves of the formula.
	Poly-log time uniformity is also preserved. 
	Therefore, we can use $(c+o(1)) \log n$-depth NAND formulas for \SAT{} on $n^{1+o(1)}$
	inputs to decide $L$ by composing them with the first-order projection; 
	the resulting formula has size $(c+o(1)) \log n^{1+o(1)} = (c+o(1)) \log n$,
	therefore $L\in\NDL{c}$.

	We now use a padding argument to lift this result to $\NTIME[n^d]$.
	Let $d>1$, let $L\in \NTIME[n^d]$ and let $M$ be a nondeterministic machine that decides $L$
	in time $n^{d+o(1)}$.
	Consider the padded language $L' = \{x1^{|x|^d - |x|} \mid x\in L\}$.
	Given an input $y$ of length $m$, we can decide whether $y\in L'$ in time 
	$m^{1+o(1)}$ as follows:
	\begin{itemize}
		\item if $y$ is not of the form $x1^{|x|^d - |x|}$ for some $x$, reject $y$.
		This can be done in time $O(m)$.
		\item otherwise, $y = x1^{|x|^d - |x|}$. Run $M$ on $x$, and accept $y$ 
		if and only it accepts $x$. This takes time $|x|^{d+o(1)} = m^{1+o(1)}$.
	\end{itemize}
	This implies $L'$ is in $\NTIME[n]$, which in turn implies that $L'\in\NDL{c}$.
	We can then use the formula descriptor $A$ for $L'$ 
	to describe formulas of depth $(c + o(1))\log \left(n^{d+o(1)}\right) = (c\cdot d + o(1))\log n$ for $L$.
	On input $x,b,n$, the descriptor machine $B$ for $L$  
	calls $A$ on $x, b, n^d$.
	If $A$ outputs a gate type or a constant, $B$ outputs the same information.
	If $A$ outputs a literal $x_i^*$, $B$ outputs $x_i^*$ if $i \le n$, and 1 otherwise.
\end{proof}

@@

%%%%%%%%%%%%%%%%%%%%%%
\subsection{Conditional Depth hierarchy theorem for \textsf{NAND} formulas}

We now prove the Conditional Depth hierarchy theorem, 
stated without proof in Section~\ref{sec:overview}.
\begin{theorem}
	If there exists $c > 0$ such that $\SAT \subseteq \NDL{c}$, 
	then for all $b > a > 0$, we have
	\[\NDL{a} \subsetneq \NDL{b}.\] 
\end{theorem}
\begin{proof}

\end{proof}
